Supplementary material for: non-coding RNAs of bird genomes

### 2014-04-07 paper/manuscript.tex
Updated manuscript and figures.

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds.  We also
demonstrate extensive conservation of classical ncRNAs (e.g., tRNAs)
and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
birds. Furthermore, we have discovered apparent ``losses'' in several
RNA families, these illustrate the complexity of bird genomes. In
particular, issues with assembling microchromosomes using
next-generation sequencing methods. These combined results illustrate
the utility of applying homology based methods for annotating
vertebrate genomes and illustrate many complex evolutionary patterns
within the avian ncRNA cohort.
\end{abstract}

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}.  {\bf Using accurate homology
  search tools and evolutionary constraints, we produce an accurate
  set of genome annotations for 48 bird species for ncRNAs that are
  conserved across the avian species. This conservative set of
  annotations is expected to contain the core avian ncRNAs.}
We focus our
report on the unusual results within the avian lineages. These are
either unexpectedly well-conserved RNAs or unexpectedly
poorly-conserved RNAs. The former are RNAs we would not have expected
to be conserved between the birds and the organisms these genes were
initially identified in; Usually, this hypothesis is based upon the
function of the RNA which is not conserved in avian species. The
latter are apparent losses of RNA genes that were expected to be
conserved; usually, this hypothesis is based upon the conservation of
these RNAs in other vertebrate species.

Here, we consider three categories of apparent loss: First, genuine
gene losses in the avian lineage. Second, ``divergence'' where RNA genes
have undergone such significant sequence and structural alternations
that homology search tools can no longer detect a relationship between
other vertebrate exemplars and avian varieties. Third, ``missing'' genes
that failed to be captured in the available, largely
fragmented, avian genomes. We postulate that the latter
category is likely to be prevalent in comparative avian genome studies
given the distinctive organisation of the avian genome. Namely, the
avian karyotype is characterised by a large number of chromosomes
(average $2n \approx 80$) generally consisting of a few larger
``macrochromosomes'' and many smaller ``microchromosomes'' \cite{Griffin:2007}. This 'so many, so small' pattern presents significant
assembly challenges \cite{Ellegren:2005}. Indeed, of the 48 published avian
genomes, 20 of which are high-coverage ($>50X$), only two are
chromosomally assembled (chicken and zebra finch; \cite{birds:14}).

%% {\bf REWRITE: We use three
%%   models of ``loss'' that explain the data: Firstly, these could be
%%   genuine cases of an ancestral gene-loss along the avian
%%   lineage. Secondly, this could be a case of ``divergence'' where a
%%   RNA gene has undergone significant sequence and structural
%%   alterations, so much so, that our homology search tools can no
%%   longer detect a relationship between vertebrate exemplars and the
%%   avian varieties. Thirdly, we consider the possibility that the
%%   available genome assemblies have independently failed to capture
%%   these genes.}

@@

\subsection*{RNA losses, divergence or missing data?}
Much of the number of apparent losses and reduction in genomic
sequence has been extensively discussed elsewhere
\cite{Organ:2007}. Unsurprisingly, this reduction is reflected in the
copy-number of RNA genes. Some of the most dramatic examples are the
transfer RNAs and pseudogenes which average $\sim900$ and $\sim580$
copies in the human, turtle and alligator genomes, the average copies
numbers of these drop to $\sim280$ and $\sim100$ copies in the avian
genomes.

The absence of seven well-conserved ncRNA families from many or even
most bird genomes is unlikely to represent true gene losses. This
concerns in particular the telomerase RNA, the RNA components of RNase
P and MRP, the minor spliceosomal snRNAs U4atac and U11, the
selenocystein tRNA (tRNA-Sec) as well as the vault RNAs (See
Figure~\ref{fig:1}).

{\bf Microchromosome background...}

We propose three possible models of ``loss'' to explain the data:
Firstly, these could be genuine cases of an ancestral gene-loss along
the avian lineage. Secondly, this could be a case of ``divergence''
where a RNA gene has undergone significant sequence and structural
alterations, so much so, that homology search tools no longer
detect a relationship between vertebrate exemplars and the avian
varieties. Thirdly, we consider the possibility that the available
genome assemblies have independently failed to capture these genes.

The third model of loss is the most likely explanation. These families
range in conservation level, from being ubiquitous to cellular-life
(RNase P RNA and tRNA-Sec), to present in the majority of eukaryotes
(RNase MRP, U4atac and U11) to present in most Bilateria
(Vault). Therefore, the loss or even diversification of these families
in the avian lineage is unlikely.

One defining feature of the avian genomes is the presence of
``microchromosomes''. As mentioned above, these are often difficult to
assemble. The most complete available avian genomes are the chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004} and the
zebrafinch \cite{Warren:2010}. The ``lost ncRNAs'' (loncRNAs) were
found to more prevalent in the more complete genomes 6/7 in chicken
and 3/7 in the zebrafinch. We examined the distribution of these
loncRNAs and discovered that only the chicken RNase MRP gene is found
on a macrochromosome. A Fisher's exact test showed that the loncRNAs
are significantly enriched within microchromosomes ($P<10^{16}$) in
both the chicken and the zebrafinch genomes (combined $P<2\times
10^{31}$).

{\bf CUT?}

In order to get an idea to what extent the
absence of these RNAs from the \texttt{infernal-based} annotation is
caused by sequence divergence beyond the thresholds of the Rfam CMs
and/or missing or incomplete data, we complemented our analysis by
dedicated searches for a few of these RNA groups.

The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
for specificity and thus misses several occurrances that are easily
found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
sequences appear degraded at the ends, which may be explained e.g.\ by
low sequence quality at the very ends of contigs or scaffolds. A
\texttt{blastn} search also readily retrieves additional RNase P and
RNAse MRP RNAs, capturing only the best conserved regions. In many
cases these additional candidates are incomplete or contain
undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Since \texttt{blastn}
searches remained unsuccessfull we constructed a sauropsid-specific CM
for the vault RNA. In addition to the hits identified by the Rfam
model we obtained three additional homologs. Vault RNAs, with a size
of about 100 nt, exhibit conserved sequence patterns only at their
ends, with essentially unconstrained sequence in the central
part. Their identification is one of the well-know difficult problems
for homology search \cite{Stadler:09b}.

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Figure~\ref{fig:2})
strongly suggests that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

%\subsection*{Y RNA clusters}

Vertebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\cite{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within bird, we find an the conserved
Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most cases
consistent with breaks in the available sequence assemblies. In
several genomes we observe one or a few additional Y RNA homologs
unlinked to the canonical Y RNA cluster. These sequences can be
identified unambiguously as derived members of one of the three
ancestral paralog groups, they almost always fit less well to the
consensus (as measured by the CM bit score of paralog group specific
covariance models) than the paralog linked to cluster, and there is no
indication that any of these additional copies is evolutionarily
conserved over longer time scales. We therefore suggest that most or
all of these interspersed copies are in fact pseudogenes.

%\subsection*{MicroRNA clusters}

Nearly all microRNAs that are broadly conserved in fish, amphibians and
mammals are also conserved in the birds. Nevertheless, there are obvious
instances of microRNAs lost in all birds. For example, mammalian and
ambhibian genomes contain three loci of clustered microRNAs from the mir-17
and mir-92 families \cite{Tanzer:04}. One of these clusters (cluster II,
with families mir-106b, mir-93 and mir-25) was not found in turtles,
crocodiles and birds, see Supplemental Figure~6.

The microRNA family let-7 is the most diverse microRNA family with 14
paralogs in human. These genes also localize in 7 genomic clusters,
together with mir-100 and mir-125 miRNA families (see previous study on the
evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
we observed that cluster A - which is strongly conserved in vertebrates has
been completely lost in the avian lineage. Another obvious loss in birds is
cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
hand has been retained in all oviparous animals and completely lost later,
after the split of Theria. See Supplemental Figure~7 for details.

@@

@@

Chicken snoRNA and miRNA annotations were validated using small
RNA-seq data comprising 27 samples from 14 diferent tissues. Since not
all RNAs are expressed at any given time we expect to verify only a
fraction of annotated RNAs using the expression data. Nevertheless 242
of 691 (35\%) miRNAs as well as 328 of 376 (87\%) snoRNAs of our
homology-based annotations were found to be supported by the RNA-seq
data.

@@

Thanks to @ewanbirney for the following tweet: ``@ctitusbrown
@BioMickWatson So ... missing orthologs to chicken often mean 'gene
might be on the microchromosome'''.

We thank Matt Schwatz (Harvard) and Igor Ulitsky (Weizmann Institute
of Science) for access to the RNA-seq data.

@@

 \begin{figure}[ht]
   \includegraphics[width=0.5\textwidth]{figures/figure1.pdf}
 \end{figure}

  \subsection*{Figure 2 - Identifying missing RNAs}

Additional homologs of some sparsely represented RNA families were
discovered using dedicated search strategies combined with highly
sensitive settings, synteny information, lineage-specific CMs and
subsequent manual inspection.\label{fig:2}

 \begin{figure}[ht]
   \includegraphics[width=0.5\textwidth]{figures/figure2.pdf}
 \end{figure}

@@

This table contains the total number of annotated ncRNAs from
different RNA types in human, chicken and the median number for each
of the 48 birds. The number of chicken ncRNA that show evidence for
expression is also indicated (the percentage is given in parentheses).

False positive rate less the 10\%:

\begin{tabular}{|r|r|r|r|l|}
\hline 
\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
\hline 
                &                  &                   & \multicolumn{1}{|l|}{Chicken ncRNAs } & \\
                &                  &                   & \multicolumn{1}{|l|}{confirmed with RNA-seq} & \\
\hline
Number in human & median(48 birds) & Number in chicken & max(RNA$_i$)$>13.0$ & RNA type\\
\hline
62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
356&499.5&427&272 (63.7\%) &microRNA\\ 
281&120.0&106&88 (83.0\%) &C/D box snoRNA\\ 
336&85.5&68&46 (67.6\%) &H/ACA box snoRNA\\ 
34&13.0&12&11 (91.7\%) &Small cajal body RNA\\ 
1754&48.5&71&29 (40.8\%) &Major spliceosomal RNA\\ 
58&3.0&6&2 (33.3\%) &Minor spliceosomal RNA\\ 
525&82.0&122&79 (64.8\%) &Cis-regulatory element\\ 
316&6.5&9&3 (33.3\%) &7SK RNA\\ 
1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
9&0.0&2&1 (50.0\%) &Vault RNA\\ 
892&3.0&3&2 (66.7\%) &Y RNA\\ 
1084&173.5&300&256 (85.3\%) &Transfer RNA\\ 
80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
941&3.0&4&2 (50.0\%) &SRP RNA\\ 
607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
\hline
7340&1080.0&1194&817 (68.4\%) &Total\\ 
\hline
\end{tabular}

>>>>

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds.  We also
demonstrate extensive conservation of classical ncRNAs (e.g., tRNAs)
and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
birds. Furthermore, we describe numerous ``losses'' of several RNA
families, and attribute these to genuine loss, divergence or missing
data.  In particular, issues with assembling microchromosomes using
next-generation sequencing methods. These combined results illustrate
the utility of applying homology based methods for annotating novel
vertebrate genomes and illustrate many complex evolutionary patterns
within the avian ncRNA cohort.
\end{abstract}

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}.  {\bf Using accurate homology
  search tools and evolutionary constraints, we produce an accurate
  set of genome annotations for 48 predominantly non-model bird
  species for ncRNAs that are conserved across the avian species. This
  conservative set of annotations is expected to contain the core
  avian ncRNAs.}  We focus our report on the unusual results within
the avian lineages. These are either unexpectedly well-conserved
ncRNAs or unexpectedly poorly-conserved ncRNAs. The former are ncRNAs
that are not expected to be conserved between the birds and the other
vertebrates, particularly those ncRNAs whose function is not conserved
in birds.  The latter are apparent losses of ncRNAs expected to be
conserved; Here, we consider three categories of such ``loss'': First,
genuine gene losses in the avian lineage where ncRNAs well conserved
in other vertebrates are completely absent in birds.  Second,
``divergence'' where ncRNAs have undergone such significant sequence
and structural alternations that homology search tools can no longer
detect a relationship between other vertebrate exemplars and avian
varieties.  Third, ``missing'' ncRNAs that failed to be captured in
the available, largely fragmented, avian genomes.  We postulate that
the latter category is likely to be prevalent in comparative avian
genome studies given the distinctive organisation of the avian
genome. Namely, the avian karyotype is characterised by a large number
of chromosomes (average $2n \approx 80$) generally consisting of a few
larger ``macrochromosomes'' and many smaller ``microchromosomes''
\cite{Griffin:2007}. This ’so many, so small’ pattern presents
significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
48 published avian genomes, 20 of which are high-coverage ($>50X$ ),
only two are chromosomally assembled (chicken and zebra finch;
\cite{birds:14,Warren:2010}).

@@

\subsection*{Unexpectedly poorly conserved  ncRNAs: genuine loss, divergence or missing  data?}

\subsubsection*{Genuine loss}

The overall reduction in avian genomic size has been extensively
discussed elsewhere \cite{Organ:2007}. Unsurprisingly, this reduction
is reflected in the copy-number of ncRNA genes. Some of the most
dramatic examples are the transfer RNAs and pseudogenes which average
$\sim900$ and $\sim580$ copies in the human, turtle and alligator
genomes, the average copies numbers of these drop to $\sim280$ and
$\sim100$ copies in the avian genomes. In addition to reduction in
copy-number, the complete absence of several otherwise ubiquitous
vertebrate ncRNAs in the avian lineage are indicative of genuine gene
loss.

Namely, mammalian and amphibian genomes contain three loci of
clustered microRNAs from the mir-17 and mir-92 families
\cite{Tanzer:04}. One of these clusters (cluster II, with families
mir-106b, mir-93 and mir-25) was not found in turtles, crocodiles and
birds (see Supplemental Figure~6). In addition, the microRNA family
let-7 is the most diverse microRNA family with 14 paralogs in
human. These genes also localize in 7 genomic clusters, together with
mir-100 and mir-125 miRNA families (see previous study on the
evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids we
observed that cluster A - which is strongly conserved in vertebrates
has been completely lost in the avian lineage.  Another obvious loss
in birds is cluster F, containing two let-7 microRNA paralogs. Cluster
H, on the other hand has been retained in all oviparous animals and
completely lost later, after the split of Theria (see Supplemental
Figure~7).

\subsubsection*{Divergence}

In order to determine to what extent the absence of these RNAs from
the infernal-based annotation is caused by sequence divergence beyond
the thresholds of the Rfam CMs, we complemented our analysis by
dedicated searches for a few of these RNA groups. Our ability to find
additional homologs for several RNA families that fill gaps in the
abundance matrices (Figure ?) strongly suggests that conspicuous
absences, in particular of LUCA and LECA RNAs, are caused by
incomplete data in the current assemblies and sequence divergence
rather then genuine losses.

Vertebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\cite{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within the avian lineage, we find a
conserved Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most
cases consistent with breaks in the available sequence assemblies.  In
several genomes we observe one or a few additional Y RNA homologs
unlinked to the canonical Y RNA cluster. These sequences can be
identified unambiguously as derived members of one of the three
ancestral paralog groups, they almost always fit less well to the
consensus (as measured by the CM bit score of paralog group specific
covariance models) than the paralog linked to cluster, and there is no
indication that any of these additional copies is evolutionarily
conserved over longer time scales. We therefore suggest that most or
all of these interspersed copies are in fact pseudogenes.

\subsubsection*{Missing data}

Seven families of ncRNAs were found in some avian genomes but not
others (Figure~\ref{fig:1}). These families range in conservation level
from being ubiquitous to cellular-life (RNase P and tRNA-sec), present
in most Bilateria (vault), present in the majority of eukaryotes
(RNase MRP, U4atac and U11) and present in all vertebrates
(telomerase) \cite{Hoeppner:2012}. Therefore, the genuine loss or even
diversification of these ncRNA families in the avian lineage is
unlikely. Rather, this lack of phylogenetic signal, combined with the
fragmented nature of the vast majority of these genomes described
above (i.e., of the 48 avian genomes, only the chicken and zebra finch
are chromosomally assembled \cite{birds:14,Warren:2010}), suggests the most likely
explanation is that these ncRNA families are indicative of missing
data. Indeed, of the seven missing ncRNA families, six where found in
the chicken genome and three were found in the zebra finch
genome. Furthermore, only one of these (RNase MRP) is found on a
macrochromosome, and all remaining missing ncRNAs are found on
microchromosomes (see Supplemental Table ?). A Fisher’s exact test
showed that there is significantly more missing ncRNAs on
microchromosomes than macrochromosomes ($P<10^{16}$ for both the
chicken and zebra finch). Thus, we suggest that many of
these ncRNAs families are missing because: (1) they are predominantly
found on microchromosomes [this study] and (2) the vast majority of
avian microchromosomes remain unassembled \cite{birds:14,Ellegren:2005}.

To wit, we performed dedicated searches for a selection of these
missing ncRNA families. Here, tRNAscan is tuned for specificity and
thus misses several occurrences of tRNA-sec that are easily found in
the majority of genomes by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
sequences appear degraded at the ends, which is likely due to low
sequence quality at the very ends of contigs or scaffolds. A \texttt{blastn}
search also readily retrieves additional RNase P and RNAse MRP RNAs in
the majority of genomes, albeit only the best conserved regions are
captured. In many cases these additional candidates are incomplete or
contain undetermined sequence, which explains why they are missed by
the CMs \cite{Stadler:09b,Kolbe:2009}.

@@

\section*{Experimentally confirmed ncRNAs}

The ncRNAs presented here have been identified using homology models
and are evolutionarily conserved in multiple avian species. In order
to further validate these predictions we have used RNA-seq from
multiple chicken tissues. After mapping chicken RNA-seq data to the
chicken genome (see Methods for details), we identified a threshold
for calling a gene as expressed by limiting our estimated
false-positive rate to less than 10\%. This was estimated using
randomly selected, un-annotated and presumably un-expressed regions of
the genome. Some of which may be genuinely expressed, meaning the true
FDR is potentially much lower than 10\%. The $<10\%$ FDR constraint,
corresponded to identifying at least one experimental condition where
13 RNA-seq reads or more were identified, that mapped to a gene of
interest. Overall, the number of ncRNAs we have identified in this
work that are expressed above background levels is 68.4\% (see
Table~\ref{table:1}). This shows that 6.9 times more of our ncRNAs are
expressed than expected by chance (Fisher's exact test:
$P<10^{16}$). It is likely that this is an underestimate of the
fraction of our annotations that are genuinely expressed, as only a
fraction of the developmental stages and tissues essential for forming
a chicken have been sampled with RNA-seq. Furthermore we now know that
some ncRNAs are expressed in highly specific conditions \cite{} ({\bf
  the worm miRNA in 1 cell paper}).

The classes of RNAs where the majority of our annotations were
confirmed includes microRNAs, snoRNAs, cis-regulatory elements, tRNAs,
SRP RNA and RNase P/MRP RNA. The RNA-seq data did fail to detect a
telomerase RNA transcript ({\bf ARE THESE ONLY EXPRESSED DURING
  MEIOSIS?}). The fractions for the other classes were well above the
10\% FDR threshold but admittedly were low for 7SK RNA, the minor
spliceosomal RNAs and the lncRNAs. There are a number of possible
explanations for this: the multiple copies of the 7SK RNA may be
functionally redundant and can therefore compensate for one another;
The minor spliceosome is active during ({\bf SOME OBSCURE CONDITION
  HOPEFULLY???}); and the lncRNAs are generally expressed at low
levels under specific conditions ({\bf cite RNACAPTURE and MATTICK'S
  BRAIN ATLAS PAPER}).

%totalExpressed[817] totalNegControlsExpressed[117] totalGenes[1194]

@@

{\bf Rewrite:}
Chicken snoRNA and miRNA annotations were validated using small
RNA-seq data comprising 27 samples from 14 diferent tissues. Since not
all RNAs are expressed at any given time we expect to verify only a
fraction of annotated RNAs using the expression data. Nevertheless 242
of 691 (35\%) miRNAs as well as 328 of 376 (87\%) snoRNAs of our
homology-based annotations were found to be supported by the RNA-seq
data.

{\bf WHERE IS ALL THE GENOME SEQUENCE AND RNA-SEQ DATA AVAILABLE
  FROM?}

@@

Thanks to @ewanbirney for the following timely tweet: ``So ... missing
orthologs to chicken often mean 'gene might be on the
microchromosome''.

We thank Matt Schwartz (Harvard) and Igor Ulitsky (Weizmann Institute
of Science) for access to the RNA-seq data.

@@

 \begin{figure}[ht]
   \centering
   \includegraphics[width=0.95\textwidth]{figures/figure1.pdf}
 \end{figure}

@@

This table contains the total number of annotated ncRNAs from
different RNA types in human, the median number for each of the 48
birds and chicken. The number of chicken ncRNA that show evidence for
expression is also indicated (the percentage is given in parentheses).
The threshold for determining expression was selected based upon a
false positive rate of less than 10\%.\label{table:1}

\begin{tabular}{|r|r|r|r|l|}
\hline 
\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
\hline 
Number in human & median(48 birds) & Number in chicken & Chicken ncRNAs         & RNA type\\
                &                  &                   & confirmed with RNA-seq & \\
\hline
62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
356&499.5&427&272 (63.7\%) &microRNA\\ 
281&120.0&106&88 (83.0\%) &C/D box snoRNA\\ 
336&85.5&68&46 (67.6\%) &H/ACA box snoRNA\\ 
34&13.0&12&11 (91.7\%) &Small cajal body RNA\\ 
1754&48.5&71&29 (40.8\%) &Major spliceosomal RNA\\ 
58&3.0&6&2 (33.3\%) &Minor spliceosomal RNA\\ 
525&82.0&122&79 (64.8\%) &Cis-regulatory element\\ 
316&6.5&9&3 (33.3\%) &7SK RNA\\ 
1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
9&0.0&2&1 (50.0\%) &Vault RNA\\ 
892&3.0&3&2 (66.7\%) &Y RNA\\ 
1084&173.5&300&256 (85.3\%) &Transfer RNA\\ 
80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
941&3.0&4&2 (50.0\%) &SRP RNA\\ 
607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
\hline
7340&1080.0&1194&817 (68.4\%) &Total\\ 
\hline
\end{tabular}

### 2014-04-07 paper/supplementary-material.tex
Updated manuscript and figures.

\author{
Paul P. Gardner\correspondingauthor$^{1,2}$
\email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Sarah W. Burge$^3$
\email{sb30@sanger.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Tammy E. Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
}

@@

>>>>

\author{
Paul P. Gardner\correspondingauthor$^{1,2}$
\email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Sarah W. Burge$^3$
\email{sb30@sanger.ac.uk},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Tammy E. Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
}

@@

\subsection*{Unexpectedly poorly conserved  ncRNAs: testing divergence}

In order to get an idea to what extent the
absence of these RNAs from the \texttt{infernal-based} annotation is
caused by sequence divergence beyond the thresholds of the Rfam CMs
and/or missing or incomplete data, we complemented our analysis by
dedicated searches for a few of these RNA groups.

The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
for specificity and thus misses several occurrances that are easily
found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
sequences appear degraded at the ends, which may be explained e.g.\ by
low sequence quality at the very ends of contigs or scaffolds. A
\texttt{blastn} search also readily retrieves additional RNase P and
RNAse MRP RNAs, capturing only the best conserved regions. In many
cases these additional candidates are incomplete or contain
undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Since \texttt{blastn}
searches remained unsuccessfull we constructed a sauropsid-specific CM
for the vault RNA. In addition to the hits identified by the Rfam
model we obtained three additional homologs. Vault RNAs, with a size
of about 100 nt, exhibit conserved sequence patterns only at their
ends, with essentially unconstrained sequence in the central
part. Their identification is one of the well-know difficult problems
for homology search \cite{Stadler:09b,Kolbe:2009}.

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Figure~\ref{fig:2})
strongly suggests that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

 \begin{figure}[ht]
   \includegraphics[width=0.5\textwidth]{figures/figure2.pdf}
\caption{Additional homologs of some sparsely represented RNA families were
discovered using dedicated search strategies combined with highly
sensitive settings, synteny information, lineage-specific CMs and
subsequent manual inspection.}\label{fig:2}
 \end{figure}
